---
title: "EDI Data Archive"
author: "burkeprw"
date: "April 04, 2018"
output: github_document
---

First, load packages.

```{r message=FALSE, warning=FALSE}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(lubridate))
suppressPackageStartupMessages(library(readr))
suppressPackageStartupMessages(library(forcats))
suppressPackageStartupMessages(library(camtrapR))
suppressPackageStartupMessages(library(unmarked))
suppressPackageStartupMessages(library(standardize))
```


### Step 1: Read Data in from mobile transects




```{r warning=FALSE, message=FALSE}


##Clean up Environment

```

Look at the distribution of sampling periods across time:   

```{r}



```




```{r pressure, echo=FALSE}


```


Next, set up the covariate matrices.   
- **Observation covariates:**  
      1. **Wind Speed** (EnvCan stations averaged over 4d); standardized using `standardize:scale()`    
      2. **Temperature** (EnvCan stations averaged over 4d); standardized using `standardize:scale()`   

  

### Spatial Regression Model 01: ___Myotis lucifugus___ (MYLU)

Here I run preliminary ....  


### Literature Cited



### Session Info

```{r}
sessionInfo()
```


