---
title: "Spatial Regression Analysis"
author: "burkeprw"
date: "April 04, 2018"
output: github_document
---

First, load packages.

```{r message=FALSE, warning=FALSE}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(lubridate))
suppressPackageStartupMessages(library(readr))
suppressPackageStartupMessages(library(forcats))
suppressPackageStartupMessages(library(camtrapR))
suppressPackageStartupMessages(library(unmarked))
suppressPackageStartupMessages(library(standardize))
```


### Step 1: Read Data in from mobile transects

The data for regression analysis was collected passively during mobile transects in a vehicle traveling faster than the speec of bat flight. Therefore, each observation is unique and an index of abundance.  


I read the .rds file saved in [Task 1](https://github.com/burkeprw/rsh_zool500_bats/blob/master/r_scripts/t1_tidydata/ghd01_tidydata.md). I also read data files from BC Agricultural Land Use Inventory (ALUI) for cover type and land use data. It is necessary to make the following formatting adjustments to the .csv data:  
- DateTimeOriginal column is formatted as YYYY-MM-DD hh:mm:ss in Excel before saving, then read into R using `format = "%Y-%m-%d %H:%M:%S`  


```{r warning=FALSE, message=FALSE}


##Clean up Environment

```

Look at the distribution of sampling periods across time:   

```{r}



```




```{r pressure, echo=FALSE}


```


Next, set up the covariate matrices.   
- **Observation covariates:**  
      1. **Wind Speed** (EnvCan stations averaged over 4d); standardized using `standardize:scale()`    
      2. **Temperature** (EnvCan stations averaged over 4d); standardized using `standardize:scale()`   

  

### Spatial Regression Model 01: ___Myotis lucifugus___ (MYLU)

Here I run preliminary ....  


### Literature Cited



### Session Info

```{r}
sessionInfo()
```


